\newcounter{loopcntr}
\newcommand{\rpt}[2][1]{%
  \forloop{loopcntr}{0}{\value{loopcntr}<#1}{#2}%
}
\newcommand{\on}[1][1]{
  \forloop{loopcntr}{0}{\value{loopcntr}<#1}{&\cellcolor{gray}}
}
\newcommand{\off}[1][1]{
  \forloop{loopcntr}{0}{\value{loopcntr}<#1}{&}
}
\newcommand{\onx}[1][1]{
  \forloop{loopcntr}{0}{\value{loopcntr}<#1}{&\cellcolor{cyan}}
}


\noindent\begin{tabular}{p{0.27\textwidth}*{16}{|p{0.02\textwidth}}|}
% The top line
\textbf{Gantt chart}
           & \multicolumn{7}{c|}{2018}
           & \multicolumn{9}{c|}{2019}\\
% The second line, with its five years of four quarters
% \rpt[2]{& 1 & 2 & 3 & 4} \\
& 06 & 07 & 08 & 09 & 10 & 11 & 12
& 01 & 02 & 03 & 04 & 05 & 06 & 07 & 08 & 09\\

\hline
% using the on macro to fill in twenty cells as `on'

A. Modular linearity with Capabilities  \on[3] \off[13] \\
\hline
B. Semantic approach to modular linearity \on[5] \off[11]\\
\hline
C. Paxos \& Mungo \off[5] \on[4] \off[7] \\
\hline
D. Thesis write-up  \off[6] \on[4] \off[6]\\
\hline
\end{tabular}

The last six months will be used for any task that overruns its allotted time.
\begin{enumerate}[A]
\item Modular linearity with Capabilities
Over the coming months my focus will be on finishing and publishing this work.

\item Semantic approach to modular linearity
I have already begun research on this, so this will be carried out in the next five months.
\item Paxos \& Mungo
The implementation of Paxos may need some refinement as the language in which is implemented might itself be refined. This however is expected to be a minor task. The bulk of the work remaining is exploring the extent to which this can be translated into the Mungo setup and implementing any extensions necessary to support it.
\item Thesis write-up
In the final four months outlined here I will focus exclusively on the write up of the thesis. As writing material relevant to the final thesis is a significant part of my day to day activities I do not foresee there being any issue in producing the final thesis in this timescale.
I expect the thesis will structured as following:
\input{sections/thesisoutline.tex}

\end{enumerate}

% In the final three months outlined here I will focus exclusively on the write up of the thesis. As writing material relevant to the final thesis is a significant part of my day to day activities I do not foresee there being any issue in producing the final thesis in this timescale.
