\section{Work Undertaken}
\label{Research}
%
% In the past year I have largely hit the goals agreed upon at the
% last annual review, an overview of which I present here. Some items, remained on hold, while other research problems appeared.

Over the past year the focus of my research has shifted to a more theoretical one. As such, some of the goals stated last year have been met, while others have been superseded by new ones. An overview of which I present here.

\subsection{Modular linearity}
\label{sub:modlinearity}
Existing work typically assumes linear type systems as necessary for session types. In order to maintain session fidelity and ensure that all communication actions in a session type occur, session type systems typically require that endpoints are used linearly: each endpoint must be used exactly once.
This work sets out to investigate whether linear types are necessary to provide resource control/ownership for session types; what are the characteristics necessary to assure session fidelity; what other semantic properties should/might be guaranteed; how can more general approaches to resource control/ownership be integrated with session types.

Our approach builds on work done to deal with memory management, explicit region deallocation in the Capability Calculus~\cite{Crary:1999:TMM:292540.292564}; changing the type of a mutable object whenever the contents of the object is changed~\cite{Ahmed:2007:LLL:1365997.1366003}; or aliasing data structures that point to linear objects~\cite{FahndrichM:adofpl1}. The work presented below is inspired by the work done by Fähndrich and DeLine~\cite{FahndrichM:adofpl1}


\subsubsection{Capabilities in $\pi$-calculus with sessions}

\input{sections/capabilities.tex}

An implementation of this work is planned for later in the year. It will be implemented within the Mungo tool\cite{kouzapas16}, developed at Glasgow University. This will allow for more flexible aliasing.

\subsubsection{A semantic approach to $\pi$-calculus with sessions}
\input{sections/semantics.tex}


\input{sections/paxos.tex}

\subsection{Other Activities}
\label{sec:Activities}

As part of the third year of PhD, various additional activities have been undertaken, such as attending conferences, ABCD group meetings of various sizes, seminars and talks(e.g. The Scottish programming language seminar series, FATA seminars) which improved my knowledge of the field and gave me some insight of the exciting research ongoing. Notably, I have attended the Oregon Programming Languages Summer School\footnote{https://www.cs.uoregon.edu/research/summerschool/summer17/}
which was a great opportunity to learn more about programming languages, from  foundational work on semantics and type theory to advanced program verification techniques; and had a big impact on the direction my research has taken since.

% I plan on attending the Oregon Programming Languages Summer School\footnote{https://www.cs.uoregon.edu/research/summerschool/summer17/} from 23 June to 8 July.
